---
title: "CpG Traditional Analysis"
author: "Paolo Cozzi"
date: today
toc: true
number-sections: true
cache: true
format:
  html:
    code-line-numbers: true
    html-math-method: katex
    df-print: paged
    code-fold: true
    toc-depth: 5
    embed-resources: true  # Embed all resources directly into the HTML file
knitr:
  opts_chunk:
    dpi: 100
editor: source
---

```{r setup, output=FALSE}
library(here)
library(GenomicRanges)
library(rtracklayer)
library(ggplot2)
library(hexbin)
library(Gviz)
```

Importing bed file:

```{r import-bed}
bed_file <- here::here("bed/ARS-UCD2.0_CpG-Islands_unix_buffer_merged.bed")
gpg_buffer <- rtracklayer::import.bed(bed_file)

# let's check the first few rows
head(gpg_buffer)
```

Importing bed file with methylation data:

```{r import-methylation, cache=FALSE}
# do not cache this chunk, is too big
bedmethyl_file <- here::here("output_methylong-5mC_5hmC-traditional/ont/A19_jun/pileup/A19_jun.bed.gz")
col_names <- c("chrom", "chromStart", "chromEnd", "name", "score", "strand", "thickStart", "thickEnd", "color", "valid_coverage", "percent_modified", "count_modified", "count_canonical", "count_other_mod", "count_delete", "count_fail", "count_diff", "count_nocall")
col_types <- readr::cols(
  chrom = readr::col_character(),
  chromStart = readr::col_integer(),
  chromEnd = readr::col_integer(),
  name = readr::col_character(),
  score = readr::col_integer(),
  strand = readr::col_character(),
  thickStart = readr::col_integer(),
  thickEnd = readr::col_integer(),
  color = readr::col_character(), # RGB triplet as string
  valid_coverage = readr::col_integer(),
  percent_modified = readr::col_double(),
  count_modified = readr::col_integer(),
  count_canonical = readr::col_integer(),
  count_other_mod = readr::col_integer(),
  count_delete = readr::col_integer(),
  count_fail = readr::col_integer(),
  count_diff = readr::col_integer(),
  count_nocall = readr::col_integer()
)
bedMethyl <- readr::read_tsv(bedmethyl_file, col_names = col_names, col_types = col_types, comment = "#")
bedMethyl$strand[bedMethyl$strand == "."] <- "*"
head(bedMethyl)
```

Let's filter methylation data and focus on `gpg_buffer` regions: let's transform
methylation `score` into a `GRanges` object and filter it based on the `gpg_buffer` regions.

```{r filter-methylation-data}
gr_methylation <- GRanges(
  seqnames = bedMethyl$chrom,
  ranges = IRanges(start = bedMethyl$chromStart+1, end = bedMethyl$chromEnd),
  strand = bedMethyl$strand,
  score = bedMethyl$score,
  percent_modified = bedMethyl$percent_modified
)
hits <- GenomicRanges::findOverlaps(gr_methylation, gpg_buffer)
filtered_methylation <- gr_methylation[queryHits(hits)]
head(filtered_methylation)
coverage_data <- as.data.frame(mcols(filtered_methylation))
head(coverage_data)
```

Plot coverage of methylation data in `gpg_buffer` regions:

```{r plot-coverage-boxplot}
ggplot(coverage_data, aes(y = score)) +
  geom_boxplot(fill = "skyblue") +
  scale_y_continuous(trans = "log10", limits = c(NA, 100)) +
  labs(
    title = "Distribution of Scores (Coverage) in CpG Buffer Regions",
    y = "Methylation Coverage (log10 scale)"
  ) +
  theme_minimal()
```

Plot the distribution of percent modified methylation:

```{r plot-percent-modified}
ggplot(coverage_data, aes(y = percent_modified)) +
  geom_boxplot(outlier.shape = NA, fill = "skyblue") +
  labs(
    title = "Distribution of Percent Modified Methylation",
    y = "Percent Modified Methylation (%)"
  ) +
  theme_minimal()
```

Plot a hexbin plot of percent modified methylation vs coverage:

```{r plot-percent-modified-coverage}
ggplot(coverage_data, aes(x = score, y = percent_modified)) +
  geom_hex(bins = 50) +
  scale_x_log10() +
  scale_fill_viridis_c(trans = "log10") +  # <--- Scala logaritmica sui conteggi
  labs(
    title = "Hexbin: Percent Modified vs Coverage (log10 scale)",
    x = "Coverage (log10 scale)",
    y = "Methylation (%)",
    fill = "Count (log10)"
  ) +
  theme_minimal()
```
